Do fast stellar centroiding methods saturate the Cramer-Rao lower bound?

### 2016-09-23 ms/ms.tex
polshing up the intro

Accuarate estimates of the centers of point
sources, which are convolved with telescope point spread function (and atmospheric PSF in case of
ground based telescopes), and also the pixel response function, are crucial to further steps of
astronomical image processing. For instance, proper measurement of the shapes of galaxies
requires interpolation of the PSF estimates from the positions of stars across the
image to the positions of galaxies. At the position of each star, the PSF is wrongly estimated by sub-pixel 
shifting of the star so that it is centered on its centroid. If the sub-pixel shifts are wrong, then 
the PSF estimates will be biased. Moreover, measurements of the parallaxes and the proper motions of stars
depends on how well we can measure their centroids.

@@

In this paper, we study the optimality of various techniques for centroiding 
faint, unsaturated stars. That is, we apply a number of centroiding methods 
to a large number of simulated faint stars, assuming uncorrelated Gaussian noise, 
with different signal-to-noise ratio and size realizations. Error from star 
centroiding methods, will always have a theoretically-set lower bound, 
known as the Cram\'{e}r-Rao lower bound (hereafter denoted by CRLB) which 
has an inverse relation with the signal-to-noise-ratio of stars. In the context 
of astrometry, the Cram\'{e}r-Rao lower bound saturation for least-squares 
estimators has been tested in specific limits in which the centroiding bias is 
negligible (\citealt{lobos}).

Reaching the optimal measurement of the centroids of stars however, is limited
by the lack of knowledge about the exact shape of the PSF, and also presence of noise;
both sky noise and CCD readout noise. Thus we need a fast method that returns
optimal estimates of centroids in a realistic range of signal-to-noise ratios.

To date, a number of softwares packages have been designed for the purpose of extracting astronomical
sources and making catalogs. One of these softwares is SExtractor (\citealt{sextractor}),
whose centroiding method involves first finding the zeroth moment of the object
as a first-order estimate, and then iteratively correcting the centroid by computing
the zeroth order moment of the object weighted by a Gaussian window function,
until the correction falls below a particular threshold value.
The width of the Gaussian window function is set by the object's half-light radius.
Other examples are DAOPHOT (\citealt{daophot}), and DOPHOT (\citealt{dophot})
which both assume analytic models for the stellar PSF profiles with centroid
coordinates being free parameters of these models.
DAOPHOT (DOPHOT) finds the centroids by fitting a Gaussian (a truncated power series) to 
the light profile of stars.

Given an analytic expression for the PSF model,
we derive an expression for the fundamental lower bound on the centroiding error as
a function of the parameters of the PSF model (e.g. PSF size),
and signal-to-noise-ratio of stars. We create two sets of simulations for which we 
know the CRLB, one with variable signal-to-noise ratio and constant FWHM, and one 
with variable FWHM and constant signal-to-noise ratio. After applying
different centroiding methods to the simulations,
we investigate how close these methods can get to saturating the CRLB for
various ranges of background noise level and PSF FWHM.

In this work, we focus on four centroiding methods. The first method
involves fitting a PSF profile, assuming that we have a good PSF estimate, to the star. 
The second method estimates the centroid of a star by fitting a 2d second-order polynomial to
 the 3$\times$3 patch around the brightest pixel of the image after convolution with the PSF. 
The third method centroids stars by
 smoothing the image of stars by a Gaussian kernel of a fixed size,
 and then applying the same 3$\times$3 polynomial trick to the smooth
 image. This method is fast and does not require any knowledge of the 
PSF. The last method we consider, is a center of light centroiding 
(measurement of a first moment), applied to the 
7$\times$7 patch around the brightest pixel of the image.

>>>>

Accuarate estimates of the centers of point
sources, which are convolved with telescope point spread function (and atmospheric PSF in case of
ground based telescopes), and also the pixel response function, are crucial to further steps of
astronomical image processing. For instance, proper measurement of the shapes of galaxies
requires interpolation of the PSF estimates from the positions of stars across the
image to the positions of galaxies. At the position of each star, the PSF is estimated by sub-pixel 
shifting of the star so that it is centered on its centroid. If the sub-pixel shifts are wrong, then 
the PSF estimates will be biased. Moreover, measurements of the parallaxes and the proper motions of stars
depends on how well we can measure their centroids.

@@

In this paper, we study the optimality of various techniques for centroiding 
faint, unsaturated stars. Our requirement for optimality is saturation of the Cram\'{e}r-Rao 
lower bound by a centroiding method. We apply a number of centroiding methods 
to a large number of simulated faint stars, assuming uncorrelated Gaussian noise, 
with different signal-to-noise ratio and size realizations. Error from star 
centroiding methods, will always have a theoretically-set lower bound, 
known as the Cram\'{e}r-Rao lower bound (hereafter denoted by CRLB) which 
has an inverse relation with the signal-to-noise-ratio of stars. In the context 
of astrometry, the Cram\'{e}r-Rao lower bound saturation for least-squares 
estimators has been tested in specific limits in which the centroiding bias is 
negligible (\citealt{lobos}).

Reaching the optimal measurement of the centroids of stars however, is limited
by the lack of knowledge about the exact shape of the PSF, and also presence of noise;
both sky noise and CCD readout noise. Thus we want investigate whether fast centroiding 
estimates can saturate the bound in a realistic range of signal-to-noise ratios.

%To date, a number of softwares packages have been designed for the purpose of extracting astronomical
%sources and making catalogs. One of these softwares is SExtractor (\citealt{sextractor}),
%whose centroiding method involves first finding the zeroth moment of the object
%as a first-order estimate, and then iteratively correcting the centroid by computing
%the zeroth order moment of the object weighted by a Gaussian window function,
%until the correction falls below a particular threshold value.
%The width of the Gaussian window function is set by the object's half-light radius.
%Other examples are DAOPHOT (\citealt{daophot}), and DOPHOT (\citealt{dophot})
%which both assume analytic models for the stellar PSF profiles with centroid
%coordinates being free parameters of these models.
%DAOPHOT (DOPHOT) finds the centroids by fitting a Gaussian (a truncated power series) to 
%the light profile of stars.

Given an analytic expression for the PSF model adopted in this study,
we derive an expression for the fundamental lower bound on the centroiding error as
a function of the parameters of the PSF model (e.g. PSF size),
and signal-to-noise-ratio of stars. We create two sets of simulations for which we can 
compute the CRLB, one with variable signal-to-noise ratio and constant FWHM, and one 
with variable FWHM and constant signal-to-noise ratio. After applying
different centroiding methods to the simulations,
we investigate how close these methods can get to saturating the CRLB for
various ranges of background Gaussian noise level and PSF FWHM.

In this work, we focus on four centroiding methods. The first method is the maximum-likelihood 
estimator which involves fitting a PSF profile, assuming that we have a good PSF estimate, to the star. 
The second method estimates the centroid of a star by fitting a 2d second-order polynomial to
 the 3$\times$3 patch around the brightest pixel of the image after convolution with the PSF. 
The third method centroids stars by
 smoothing the image of stars by a Gaussian kernel of a fixed size,
 and then applying the same 3$\times$3 polynomial trick to the smooth
 image. This method is fast and does not require any knowledge of the 
PSF. The last method we consider, is a center of light centroiding 
(measurement of a first moment), applied to the 
7$\times$7 patch around the brightest pixel of the image.
